---
title: "CLPTreatments_PISurveys"
output: html_document
---

```{r loading packages, results = "hide", fig.keep = "none"}
library(tidyverse)
library(arrow)
library(readxl)
library(lubridate)
library(sf)
library(here)

library(mgcv)
library(dlnm)
library(brms)

library(ggplot2)
library(ggplotplus)
library(viridis)
library(gridExtra)
library(cowplot)

library(dtwclust)
library(clusterCrit)
```


Create CLP treatment database.
```{r setting up treatment data}
all_IAPM_treatments <- read_csv(here("outputdata", "all_IAPM_treatments.csv"))


CLP_treatments <- all_IAPM_treatments %>%
 filter(species_code == "CLP") %>%
  filter(!is.na(start_date))

# #filtering out treatments where we only know the month
# CLP_treatments <- CLP_treatments %>%
#   # calculate the difference in start date to end date
#   mutate(
#     days_apart = as.numeric(difftime(end_date, start_date, units = "days"))
#   ) %>%
#   # 2. Keep only rows where they are 0 or 1 day apart (and drop NAs)
#   filter(!is.na(days_apart), days_apart <= 1) %>%
#   # 3. Remove the temporary calculation column
#   select(-days_apart)
```


Summarize survey data to Date-DOW level
Starting with db_sub_pa from 1_Loading and preparing PI data.R

```{r setting up PI survey data}
#load in PI data
db_sub_pa <- read_csv("~/IAPM Project/data/db_sub_pa.csv")

#load in lake LZ data, joing LZ acres
MN_lake_basin_littoral_zone <- read_csv(here("data", "MN_lake_basin_littoral_zone_15_ft_std.csv"))
CLP_treatments <- CLP_treatments %>%
  left_join(
    MN_lake_basin_littoral_zone %>% select(DOWLKNUM, LK_LZACRES), 
    by = c("DOW" = "DOWLKNUM")
  )

# Exploring DLNM possibilities ----

dat_dlnm <- db_sub_pa %>%
  filter(!is.na(depth_ft) & depth_ft <= 15) %>%
  mutate(SURVEY_START=as.Date(SURVEY_START),
         year=year(SURVEY_START),
         month=month(SURVEY_START),
         doy=yday(SURVEY_START),
         taxa_rich=rowSums(across(ceratophyllum_demersum:vallisneria_sp)),
         vegetated=ifelse(taxa_rich==0, 0, 1),
         EWM=apply(across(c(myriophyllum_spicatum, myriophyllum_spicatum_x_sibiricum)), MARGIN=1, FUN=max),
         native_rich=rowSums(across(c(ceratophyllum_demersum, 
                                      elodea_canadensis:eleocharis_acicularis,
                                      myriophyllum_sibiricum:potamogeton_robbinsii, 
                                      nitella_sp:potamogeton_strictifolius,
                                      fontinalis_sp:drepanocladus_sp,
                                      nitella_furcata:vallisneria_sp))),
         native_vegetated=ifelse(taxa_rich==0, 0, 1)) %>%
  group_by(DOW, lake_name, SURVEY_START, year, month, doy) %>%
  summarize_at(vars(ceratophyllum_demersum:vallisneria_sp, taxa_rich:native_vegetated), mean) %>%
  ungroup() %>%
  mutate(survey_taxa_rich=rowSums(across(ceratophyllum_demersum:vallisneria_sp)>0),
         survey_native_rich=rowSums(across(c(ceratophyllum_demersum, 
                                             elodea_canadensis:eleocharis_acicularis,
                                             myriophyllum_sibiricum:potamogeton_robbinsii, 
                                             nitella_sp:potamogeton_strictifolius,
                                             fontinalis_sp:drepanocladus_sp,
                                             nitella_furcata:vallisneria_sp))>0))

#decided not to remove NAs. Need to investigate this more

#join LZ acres to dat_dlnm
dat_dlnm <- dat_dlnm %>%
  left_join(
    MN_lake_basin_littoral_zone %>% select(DOWLKNUM, LK_LZACRES), 
    by = c("DOW" = "DOWLKNUM")
  )
  


#generates a list of CLP lakes
clp_lakes_dat_dlnm <- as.character((dat_dlnm %>%
                                      group_by(DOW) %>%
                                      summarize(potamogeton_crispus=max(potamogeton_crispus)) %>%
                                      filter(potamogeton_crispus > 0))$DOW)


#create unique IDs for PI surveys
dat_dlnm <- dat_dlnm %>%
  mutate(
    survey_id = row_number() 
  )

dat_dlnm <- dat_dlnm %>%
  relocate(survey_id, .before = 1)

```


Create the treatment history matrix
```{r treatment history matrix}
# Initialize matrix with 11 columns (survey_id + lag0_before + lag0_after + lag1..lag8)
treatment_history_matrix <- matrix(0, nrow = nrow(dat_dlnm), ncol = 11) %>%
  as.data.frame()

colnames(treatment_history_matrix) <- c(
  "survey_id", 
  "lag0_before", 
  "lag0_after", 
  paste0("lag", 1:8)
)

treatment_history_matrix$survey_id <- dat_dlnm$survey_id
treatment_history_matrix$SURVEY_START <- dat_dlnm$SURVEY_START
treatment_history_matrix$DOW <- dat_dlnm$DOW
treatment_history_matrix <- treatment_history_matrix %>%
  relocate(DOW, SURVEY_START, .before = 2)

# Tracking vector
target_string <- rep(NA_character_, nrow(dat_dlnm))


#Populate matrix row-by-row
for (i in 1:nrow(dat_dlnm)) {
  
  current_dow  <- dat_dlnm$DOW[i]
  survey_dt    <- dat_dlnm$SURVEY_START[i]
  survey_yr    <- dat_dlnm$year[i]
  
  # Check if the lake exists in the CLP treatment database
  if (current_dow %in% CLP_treatments$DOW) {
    
    # Subset treatment data for this specific lake
    CLP_treatments_i <- CLP_treatments %>% 
      filter(DOW == current_dow)
    
    # Track species code if present
    if ("species_code" %in% names(CLP_treatments_i)) {
      target_string[i] <- paste(unique(CLP_treatments_i$species_code), collapse = ",")
    }
    
    # --- Same-Year Logic (lag0_before vs lag0_after) ---
    same_year_tx <- CLP_treatments_i %>% filter(treatment_year == survey_yr)
    
    if (nrow(same_year_tx) > 0) {
      if (any(same_year_tx$start_date <= survey_dt)) {
        treatment_history_matrix[i, "lag0_before"] <- 1
      }
      if (any(same_year_tx$start_date > survey_dt)) {
        treatment_history_matrix[i, "lag0_after"] <- 1
      }
    }
    
    # --- Lag Years 1 through 8 ---
    for (j in 1:8) {
      target_year <- survey_yr - j
      col_name <- paste0("lag", j)
      
      if (target_year %in% CLP_treatments_i$treatment_year) {
        treatment_history_matrix[i, col_name] <- 1
      }
    }
    
  } else {
    # No treatment history for this lake DOW
    target_string[i] <- "none"
  }
}

```



```{r}
#check the number of rows of the matrix with at least one lag treatment
sum(rowSums(treatment_history_matrix[, 4:13] == 1) > 0)

#filter dat_dlnm to surveys that have previous treatments
survey_ids_treated <- treatment_history_matrix %>% #create list of surveys with any previous treatments
  filter(if_any(starts_with("lag"), ~ .x == 1)) %>%
  pull(survey_id)

dat_dlnm_surveys_w_treatments <- dat_dlnm %>% #filter
  filter(survey_id %in% survey_ids_treated)

# Verify the new row count
nrow(dat_dlnm_surveys_w_treatments)


```


For treatments in all_IAPM_treatments where we only know the month, start_date is first of the month and end_date is last of the month. This code is to filter out surveys that occur in the same month as one of these treatments where we only know the month. Only matters for lag0. I am aware this may also filter out the few lag0 treatments that happen to be on the first of the month. Will fix later.
```{r deal w lag0 treatments where we only know the month}
dat_dlnm_surveys_w_treatments <- dat_dlnm_surveys_w_treatments %>%
  # 1. Join with CLP_treatments matching on DOW and year
  left_join(
    CLP_treatments %>%
      select(DOW, treatment_year, same_year_treatment = start_date) %>%
      distinct(), # Prevents duplicates if a DOW was treated multiple times in one year
    by = c("DOW", "year" = "treatment_year")
  )

 dat_dlnm_surveys_w_treatments <- dat_dlnm_surveys_w_treatments %>% 
   relocate(same_year_treatment, .before = 5) #rearrange columns
 
 same_month_check <- dat_dlnm_surveys_w_treatments %>%
  filter(
    year(SURVEY_START) == year(same_year_treatment),
    month(SURVEY_START) == month(same_year_treatment)
  ) %>%
  filter(format(same_year_treatment, "%d") == "01")
 
 
 #filter out surveys in same_month_check from treatment matrix
 treatment_history_matrix <- treatment_history_matrix %>%
  filter(!survey_id %in% same_month_check$survey_id)
 
 survey_ids_treated <- treatment_history_matrix %>% #create list of surveys with any previous treatments
  filter(if_any(starts_with("lag"), ~ .x == 1)) %>%
  pull(survey_id)

dat_dlnm_surveys_w_treatments <- dat_dlnm %>% #filter
  filter(survey_id %in% survey_ids_treated)
 
 #check to make sure number of rows are the same
 sum(rowSums(treatment_history_matrix[, 4:13] == 1) > 0)
 nrow(dat_dlnm_surveys_w_treatments)
 
 #generate list of DOWs for lakes we have survey data and CLP treatment history
lakes_CLP_trt_survey <- dat_dlnm_surveys_w_treatments %>%
  distinct(DOW) %>%
  pull(DOW)
```



Creating the same matrix but with permit data to cross check and change 0s in original treatment history matrix that have permitted treatments. 
```{r false 0s check - permit matrix}
permits <- permits <- read.csv("~/IAPM Project/Data/apm_iapm_permit_detail.csv",
                               na.strings=c("NA", "NULL"))
lake_basin_morphology_df <- read_csv("~/IAPM Project/MAISRC-IAPM/data/MN_lake_basin_littoral_zone_15_ft_std.csv")

#setting up IAPM permit data
permits <- permits %>%
  dplyr::select(permit_number, permit_type, 
                water_resource_names, water_resource_numbers,
                treatment_method, species,
                permit_approved_acres_1997:permit_approved_acres_2025)

permits[, 7:35][is.na(permits[, 7:35])] <- 0

iapm_permits_long <- permits

colnames(iapm_permits_long)[7:35] <- 1997:2025

iapm_permits_long <- iapm_permits_long %>%
  filter(permit_type=="Invasive Aquatic Plant Management") %>%
  pivot_longer(cols=`1997`:`2025`, 
               names_to="year",
               values_to="approved_acres") %>%
  mutate(year=as.numeric(year))

CLP_permits_long <- iapm_permits_long %>%
 filter(species == "Curly-leaf Pondweed")
#rename columns
CLP_permits_long <- CLP_permits_long %>% 
  rename(DOW = water_resource_numbers) %>%
  filter(approved_acres > 0) %>%
  


#permit treatment history matrix
permit_treatment_history_matrix <- matrix(0, nrow = nrow(dat_dlnm), ncol = 10) %>%
  as.data.frame()

colnames(permit_treatment_history_matrix) <- c(
  "survey_id", 
  paste0("lag", 0:8)
)

permit_treatment_history_matrix$survey_id <- dat_dlnm$survey_id
permit_treatment_history_matrix$SURVEY_START <- dat_dlnm$SURVEY_START
permit_treatment_history_matrix$DOW <- dat_dlnm$DOW
permit_treatment_history_matrix <- permit_treatment_history_matrix %>%
  relocate(DOW, SURVEY_START, .before = 2)

# Tracking vector
target_string <- rep(NA_character_, nrow(dat_dlnm))


#Populate matrix row-by-row
for (i in 1:nrow(dat_dlnm)) {
  
  current_dow  <- dat_dlnm$DOW[i]
  survey_dt    <- dat_dlnm$SURVEY_START[i]
  survey_yr    <- dat_dlnm$year[i]
  
  # Check if the lake exists in the CLP treatment database
  if (current_dow %in% CLP_permits_long$DOW) {
    
    # Subset treatment data for this specific lake
    CLP_permits_long_i <- CLP_permits_long %>% 
      filter(DOW == current_dow)
    
    # Track species code if present
    if ("species" %in% names(CLP_permits_long_i)) {
      target_string[i] <- paste(unique(CLP_permits_long_i$species), collapse = ",")
    }
    
    # --- Same-Year Logic (lag0) ---
    same_year_tx <- CLP_permits_long_i %>% 
      filter(
        year == survey_yr,
        !is.na(approved_acres),
        approved_acres > 0
      )
    
    permit_treatment_history_matrix[i, "lag0"] <- ifelse(nrow(same_year_tx) > 0, 1, 0)
    
    # --- Lag Years 1 through 8 ---
    for (j in 1:8) {
      target_year <- survey_yr - j
      col_name <- paste0("lag", j)
      
      has_treated_lag <- CLP_permits_long_i %>% 
        filter(
          year == target_year,
          !is.na(approved_acres),
          approved_acres > 0
        ) %>% 
        nrow() > 0
      
      permit_treatment_history_matrix[i, col_name] <- ifelse(has_treated_lag, 1, 0)
    }
    
  } else {
    # No treatment history for this lake
    target_string[i] <- "none"
    
    # Explicitly ensure lag columns stay 0
    permit_treatment_history_matrix[i, paste0("lag", 0:8)] <- 0
  }
}

#check the number of surveys that had treatment histories in original matrix
sum(permit_treatment_history_matrix$survey_id %in% survey_ids_treated, na.rm = TRUE)
#check the number of surveys with permit history (at least one 1 in lag matrix)
sum(rowSums(permit_treatment_history_matrix[, 4:12] == 1) > 0)


filtered_permit_matrix <- permit_treatment_history_matrix %>%
  filter(survey_id %in% survey_ids_treated)

filtered_treatment_matrix <- treatment_history_matrix %>%
  filter(survey_id %in% survey_ids_treated)


#check how many 1s there are in each of the filtered matrices
sum(filtered_permit_matrix[, paste0("lag", 0:8)] == 1, na.rm = TRUE)
sum(filtered_treatment_matrix[, 4:13] == 1, na.rm = TRUE)


#check for any 0s in permit matrix that are 1s in treatment matrix (should be o)
# Add temp_col as column 4 temporarily
filtered_permit_matrix <- filtered_permit_matrix %>% 
  mutate(temp_col = 0, .before = 4)

#Count the mismatches
mismatch_count <- sum(
  filtered_permit_matrix[, 6:13] == 0 & filtered_treatment_matrix[, 6:13] == 1,
  na.rm = TRUE
)
mismatch_count

#Create a logical matrix of where mismatches occur (cols 5-13)
mismatch_matrix <- filtered_permit_matrix[, 6:13] == 0 & filtered_treatment_matrix[, 6:13] == 1
#Total count of mismatched cells
mismatch_count <- sum(mismatch_matrix, na.rm = TRUE)
cat("Total mismatched cells:", mismatch_count, "\n\n")
# 4. Extract rows where at least one mismatch occurred in cols 5-13
mismatch_row_indices <- which(rowSums(mismatch_matrix, na.rm = TRUE) > 0)
# 5. Get unique survey_ids for those rows
mismatched_survey_ids <- filtered_permit_matrix$survey_id[mismatch_row_indices]
# Display the survey_ids
mismatched_survey_ids


#and the other way around. This should be more than 0
mismatch_count2 <- sum(
  filtered_permit_matrix[, 6:13] == 1 & filtered_treatment_matrix[, 6:13] == 0,
  na.rm = TRUE
)

mismatch_count2

```



```{r combine treatment and permit matrices}
# 1. Identify logical condition (Permit == 1 & Treatment == 0) for cols 6 to 13
mismatch_mask <- filtered_permit_matrix[, 6:13] == 1 & filtered_treatment_matrix[, 6:13] == 0

# 2. Update filtered_treatment_matrix cells to 2 where condition is TRUE
filtered_treatment_matrix[, 6:13][mismatch_mask] <- 2

#check to make sure there are now 0 mismatched columns
mismatch_count2 <- sum(
  filtered_permit_matrix[, 6:13] == 1 & filtered_treatment_matrix[, 6:13] == 0,
  na.rm = TRUE
)
mismatch_count2


#remove temporary column
filtered_permit_matrix <- filtered_permit_matrix %>% 
  select(-temp_col)

```




```{r write csv}
write.csv(filtered_treatment_matrix, file = "~/IAPM Project/MAISRC-IAPM/analysis/PI Surveys CLP Treatment History - DLNM/filtered_treatment_history_matrix.csv")

```
