---
title: "PARsClean"
output: html_document
---
# Taken from Maddie's IAPM_PARs_clean code, but with keeping records that don't have a permit number
# last edited 5/18/2026
# this script cleans up the PARs data as downloaded from sharepoint.
# the end result is lumped treatments to the permit-year-treatment-chemical level
# there can be two rows of the same permit-year-day if there are multiple chemicals
```{r}
library(tidyverse)
library(lubridate)
library(DescTools)

ifelse(!dir.exists("data"), 
       dir.create("data"), 
       "Folder exists already")
ifelse(!dir.exists("outputdata"), 
       dir.create("outputdata"), 
       "Folder exists already")
ifelse(!dir.exists("draftcode"), 
       dir.create("draftcode"), 
       "Folder exists already")
PARs <- read_excel("~/IAPM Project/PARs-query-FINAL_exported5-27-2026.xlsx")

```

Remove columns we don't care about.
```{r}
PARs_simple <- subset(PARs, select = c(Permit_number, App_year, species_code, Lake_parent_name, k_Lake_ID, County_name, Pesticide_trade_name, Supplier, Applicator_company_name, PAR_start_date, PAR_end_date, treatment_size))
```


grouping by
```{r}
PARs_treatment_group <- PARs_simple %>% 
  
  arrange(Permit_number, Lake_parent_name, k_Lake_ID, App_year, species_code, Pesticide_trade_name, PAR_start_date) %>%
  
  # create treatment groups based on permit-year-pesticides that have close application dates and can be assumed to be the same treatment.
  ### should I add to this code so treatment group is numbered for easy grouping again
  
  group_by(Permit_number, County_name, Lake_parent_name, k_Lake_ID, App_year, species_code, Pesticide_trade_name) %>%
  mutate(
    day_gap = as.numeric(PAR_start_date - lag(PAR_start_date)),
    new_treatment = if_else(is.na(day_gap) | day_gap > 5, 1, 0), # assumes 5 day treatment window is the max.
    treatment_group = cumsum(new_treatment)
  ) %>%
   summarise(
    start_date = if(all(is.na(PAR_start_date))) {
      as.Date(NA)
      } else {
        min(PAR_start_date, na.rm = TRUE)
        },
    end_date   = if(all(is.na(PAR_end_date))) {
      as.Date(NA)
    } else {
      max(PAR_start_date, na.rm = TRUE)
    },
    treatment_size = sum(treatment_size, na.rm = TRUE))
```

#Deleting duplicates but for some reason this doesn't delete all duplicates. Need to check
```{r}
PARs_treatment_group <- PARs_treatment_group %>%
  distinct(Lake_parent_name, start_date, treatment_size, .keep_all = TRUE)
```



```{r}
write.csv(PARs_treatment_group,
          "PARs_clean1.csv",
          row.names = FALSE)
```

